Knowledge \& the web: exercise session 6

### 2015-11-22 exercise_6/ex6.tex
Decision tree part added

Apart from rule set learning, we also experimented witch decision trees. First we applied the standard decision tree in RapidMiner, it created a too complicated decision tree, as you can see in figure 1.

\begin{figure}[H]
  \centering
  \includegraphics[width = 1.5\textwidth]{decisiontreetitanic_original.pdf}
  \caption{Original decision tree with maximal depth of 20}
\end{figure}
Next we limited the maximal depth of the decision tree to 3. This gave us figure 2, which is much more usefull.
\begin{figure}[H]
  \centering
  \includegraphics[width = 1\textwidth]{decisiontreetitanic.pdf}
  \caption{Decision tree with maximal depth of 3}
\end{figure}

>>>>

Apart from rule set learning, we also experimented witch decision trees. Since we are working with a mix of nominal and numeric data, this is a suitable model, because each node can use its own \emph{rule} for splitting on an attribute. We will describe how we tweaked the learning parameters in order to achieve better results. The initial tree is shown in Figure~\ref{fig:tree_initial}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width = .6\textwidth]{tree_initial}
  \captionsetup{width=.8\textwidth}
  \caption{Tree learned with default parameters}
  \label{fig:tree_initial}
\end{figure}

\par The most obvious observation about this initial model is that female passengers with third class tickets are not yet classified very accurately, yet the tree did not split on it. In order to improve on this, we can adjust the splitting criterion to \emph{accuracy}. It will split nodes such that the accuracy of the tree increases. This results in a very sparse and large tree, which is not shown for exactly that reason. By changing the maximal depth of the tree, we can prevent it from getting too large. A maximal depth of 4 was found to provide good results. The result is shown in Figure~\ref{fig:tree_accuracy}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width = .6\textwidth]{tree_accuracy}
  \captionsetup{width=.8\textwidth}
  \caption{Tree learned with \emph{accuracy} splitting criterion and maximum depth of 4.}
  \label{fig:tree_accuracy}
\end{figure}
As we can see, this achieved the desired result of splitting the undecided node in four new nodes, three of which perform clearly better. The final step is improving prediction for males. While its accuracy is quite high, the recall for surviving males is zero. This probably happens because the gain in criterion is not sufficiently high for the pre-pruning step to split the node. By lowering the minimum gain during pre-pruning, we enable the algorithm to split easier on nodes. Because the \texttt{Fare} node is already split until max depth, we expect that only the rightmost node (males) will be further split by this setting. Figure~\ref{fig:tree_final} confirms this assumption and shows the final model. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width = .8\textwidth]{tree_final}
  \captionsetup{width=.8\textwidth}
  \caption{Tree learned with \emph{accuracy} splitting criterion, maximum depth of 4 and minimum information gain of 0.01.}
  \label{fig:tree_final}
\end{figure}
\par This model is pretty intuitive. For females, class is the most important deciding factor. Amongst the lowest class women, the ones that payed most for their ticket surprisingly have the lowest survival rate. For males, age is the most important deciding factor. Children have a far higher survival rate than men.
